%---------------------------------------------------------------------------%
%->> Frontmatter
%---------------------------------------------------------------------------%
%-
%-> 生成封面
%-
\maketitle% 生成中文封面
{\hfuzz=10pt\MAKETITLE}% 生成英文封面
%-
%-> 作者声明
%-
\makedeclaration% 生成声明页
%-
%-> 中文摘要
%-
\intobmk\chapter*{摘要}% 显示在书签但不显示在目录
\setcounter{page}{1}% 开始页码
\pagenumbering{Roman}% 页码符号

大语言模型在数学推理等可验证任务上的后训练，越来越依赖基于组采样的强化学习方法。然而，固定小组采样会导致一个经常被忽视的统计问题：当同一 prompt 采样得到的回答全对或全错时，组内奖励方差退化，进而造成零梯度更新。本文将这一现象概括为\textit{signal collapse}，并指出其根源并不完全在于题目是否可学，而在于训练预算与 prompt 难度之间存在显著错配。换言之，许多“学不到”的表面现象，实际上是固定预算在异质难度分布上无法稳定暴露正负对比信号的结果。

围绕这一问题，本文构建了一个“统计分析—采样机制—系统友好性”相统一的研究框架。基于 \texttt{Qwen2.5-Math-1.5B base} 在 \texttt{OpenR1-Math-220k} 上的 prompt 级重复采样统计与训练 checkpoint 切片，本文系统分析了固定预算、自适应固定块预算和几何块预算的差异，并刻画训练数据内部显著存在的难度长尾与分层结构。研究表明：即使组大小提升到 $32$，训练前后仍分别有 35.36\% 与 26.41\% 的 prompt 发生 signal collapse；在此背景下，自适应采样较固定预算具有更低的信号缺失率，而几何块自适应策略在保持统计效果的同时，更适合与 KV Cache 复用等系统优化结合。进一步的端到端训练结果表明，本文提出的 H-GAS 在多种基座模型与数学推理基准上均取得一致提升，并在高难度基准上表现出更明显优势；受控实验与硬件画像则显示，正样本优先退出、几何块组织、动态温度与 TSIS 共同实现了更优的预算效率、较低的 Prefill 占比与更短的统计训练时间。最后，本文进一步讨论几何块批次组织与前缀复用、在线流水线调度以及温度缩放重要性校正之间的衔接方式，从而把采样问题扩展为统计效率与系统效率的联合设计问题。

本文的工作说明，在 reasoning 强化学习后训练中，采样预算本身就是训练信号设计的一部分。与其仅从损失函数角度理解训练稳定性，不如将 prompt 级预算分配、退出规则和系统批次组织作为统一问题加以设计。本文最终主张：采样不应被视为训练外部的附属开销，而应被视为决定训练信号是否能够被有效观测的内部机制。

\keywords{大语言模型，强化学习后训练，自适应采样，信号坍塌，数学推理}% 中文关键词
%-
%-> 英文摘要
%-
\intobmk\chapter*{Abstract}% 显示在书签但不显示在目录

Reinforcement learning post-training for large language models has recently become a major approach for improving mathematical reasoning. A practical difficulty of group-based optimization, however, is that fixed small-group sampling often yields all-correct or all-incorrect response sets for the same prompt, causing the within-group reward variance to vanish and the gradient signal to collapse. This thesis studies the problem from the perspective of prompt-level sampling allocation and argues that many zero-signal cases are induced by budget mismatch rather than the intrinsic unlearnability of prompts.

To address this issue, the thesis develops a unified perspective that connects statistical analysis, sampling design, and systems-friendly execution. Using prompt-level repeated sampling statistics and checkpoint slices from the Qwen2.5-Math-1.5B base model trained on OpenR1-Math-220k, it characterizes the long-tail structure of prompt difficulty, analyzes the evolution of signal collapse during training, and compares fixed-budget, fixed-block adaptive, and geometric-block adaptive sampling strategies. The results show that the sampled training data exhibits a pronounced long-tail difficulty structure; signal collapse remains substantial even with group size $32$; adaptive sampling consistently outperforms comparable fixed budgets; and geometric-block scheduling achieves nearly the same statistical effectiveness as fixed-block scheduling while offering a more system-friendly batch structure. End-to-end RL post-training experiments further show that H-GAS delivers consistent gains across multiple base models and mathematical reasoning benchmarks, while controlled efficiency studies indicate better budget allocation, lower prefill overhead, and shorter wall-clock training time. Finally, the thesis discusses how geometric-block scheduling can better interface with prefix reuse, KV-cache-aware execution, and temperature-based importance correction in future system implementations.

These findings suggest that sampling allocation should be viewed as an integral part of signal design in reasoning-oriented reinforcement learning, together with the stopping rule and systems-level batch organization.

\KEYWORDS{large language models, reinforcement learning post-training, adaptive sampling, signal collapse, mathematical reasoning}% 英文关键词
%---------------------------------------------------------------------------%
